Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections

### 2023-05-21 jss/cheem_paper.Rmd
added ExplainPrediction and flashlight references

---
documentclass: jss
author:
  - name: Nicholas Spyrison
    orcid: 0000-0002-8417-0212
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Dept of Human Centred Computing
      | Faculty of Information Technology
    email: \email{spyrison@gmail.com}
    url: https://nspyrison.netlify.app
  - name: Dianne Cook
    orcid: 0000-0002-3813-7155
    address: |
      | Dept of Econometrics and Business Statistics
    affiliation: Monash University
  - name: Przemyslaw Biecek
    orcid: 0000-0001-8423-1823
    affiliation: |
      | Warsaw University of 
      | Technology and University
title:
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain:     "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  # For running headers, if needed
  short:     "Exploring Local Explanations"
abstract: >
    The increased predictive power of machine learning models comes at the cost of increased complexity and loss of interpretability, particularly in comparison to parametric statistical models. This trade-off has led to the emergence of eXplainable AI (XAI) which provides methods, such as local explanations (LEs) and local variable attributions (LVAs), to shed light on how a model use predictors to arrive at a prediction. These provide a point estimate of the linear variable importance in the vicinity of a single observation. However, LVAs tend not to effectively handle association between predictors. To understand how the interaction between predictors affects the variable importance estimate, we can convert LVAs into linear projections and use the radial tour. This is also useful for learning how a model has made a mistake, or the effect of outliers, or clustering of observations.  The approach is llustrated with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain:     [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: paper.bib
---

@@

There are numerous R packages currently available which provide functions for computing SHAP values, including `fastshap` [@fastshap], `kernelshap` [@kernelshap], `shapr` [@shapr], `shapviz` [@shapviz], `PPtreeregViz` [@PPtreeregViz], and the package `DALEX` has many resources [@biecek_dalex_2018]. There are many more packages only available through Github, like `treeshap` [kominsarczyk_treeshap_2021] that is used for this work. @molnar2022 provides good explanations of the different methods and how to apply them to different models.

>>>>

---
documentclass: jss
author:
  - name: Nicholas Spyrison
    orcid: 0000-0002-8417-0212
    affiliation: Monash University
    # use this syntax to add text on several lines
    address: |
      | Dept of Human Centred Computing
      | Faculty of Information Technology
    email: \email{spyrison@gmail.com}
    url: https://nspyrison.netlify.app
  - name: Dianne Cook
    orcid: 0000-0002-3813-7155
    address: |
      | Dept of Econometrics and Business Statistics
    affiliation: Monash University
  - name: Przemyslaw Biecek
    orcid: 0000-0001-8423-1823
    affiliation: |
      | University of Warsaw
      | Warsaw University of Technology
title:
  formatted: "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  plain:     "Exploring Local Explanations of Nonlinear Models Using Animated Linear Projections"
  # For running headers, if needed
  short:     "Exploring Local Explanations"
abstract: >
    The increased predictive power of machine learning models comes at the cost of increased complexity and loss of interpretability, particularly in comparison to parametric statistical models. This trade-off has led to the emergence of eXplainable AI (XAI) which provides methods, such as local explanations (LEs) and local variable attributions (LVAs), to shed light on how a model use predictors to arrive at a prediction. These provide a point estimate of the linear variable importance in the vicinity of a single observation. However, LVAs tend not to effectively handle association between predictors. To understand how the interaction between predictors affects the variable importance estimate, we can convert LVAs into linear projections and use the radial tour. This is also useful for learning how a model has made a mistake, or the effect of outliers, or clustering of observations.  The approach is llustrated with examples from categorical (penguin species, chocolate types) and quantitative (soccer/football salaries, house prices) response models. The methods are implemented in the R package cheem, available on CRAN.
keywords:
  # at least one keyword must be supplied
  formatted: [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
  plain:     [explainable artificial intelligence, nonlinear model interpretability, visual analytics, local explanations, grand tour, radial tour]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: paper.bib
---

@@

There are numerous R packages currently available which provide functions for computing SHAP values, including `fastshap` [@fastshap], `kernelshap` [@kernelshap], `shapr` [@shapr], `shapviz` [@shapviz], `PPtreeregViz` [@PPtreeregViz], `ExplainPrediction` [@ExplainPrediction], `flashlight` [@flashlight] and the package `DALEX` has many resources [@biecek_dalex_2018]. There are many more packages only available through Github, like `treeshap` [@kominsarczyk_treeshap_2021] that is used for this work. @molnar2022 provides good explanations of the different methods and how to apply them to different models.

### 2023-05-21 jss/cheem_paper.tex
added ExplainPrediction and flashlight references

\author{
Nicholas Spyrison~\orcidlink{0000-0002-8417-0212}\\Monash
University \And Dianne Cook~\orcidlink{0000-0002-3813-7155}\\Monash
University \And Przemyslaw
Biecek~\orcidlink{0000-0001-8423-1823}\\Warsaw University of\\
Technology and University
}
\title{Exploring Local Explanations of Nonlinear Models Using Animated
Linear Projections}

@@

There are numerous R packages currently available which provide
functions for computing SHAP values, including \texttt{fastshap}
\citep{fastshap}, \texttt{kernelshap} \citep{kernelshap}, \texttt{shapr}
\citep{shapr}, \texttt{shapviz} \citep{shapviz}, \texttt{PPtreeregViz}
\citep{PPtreeregViz}, and the package \texttt{DALEX} has many resources
\citep{biecek_dalex_2018}. There are many more packages only available
through Github, like \texttt{treeshap}
{[}kominsarczyk\_treeshap\_2021{]} that is used for this work.
\citet{molnar2022} provides good explanations of the different methods
and how to apply them to different models.

>>>>

\author{
Nicholas Spyrison~\orcidlink{0000-0002-8417-0212}\\Monash
University \And Dianne Cook~\orcidlink{0000-0002-3813-7155}\\Monash
University \And Przemyslaw
Biecek~\orcidlink{0000-0001-8423-1823}\\University of Warsaw\\
Warsaw University of Technology
}
\title{Exploring Local Explanations of Nonlinear Models Using Animated
Linear Projections}

@@

There are numerous R packages currently available which provide
functions for computing SHAP values, including \texttt{fastshap}
\citep{fastshap}, \texttt{kernelshap} \citep{kernelshap}, \texttt{shapr}
\citep{shapr}, \texttt{shapviz} \citep{shapviz}, \texttt{PPtreeregViz}
\citep{PPtreeregViz}, \texttt{ExplainPrediction}
\citep{ExplainPrediction}, \texttt{flashlight} \citep{flashlight} and
the package \texttt{DALEX} has many resources \citep{biecek_dalex_2018}.
There are many more packages only available through Github, like
\texttt{treeshap} \citep{kominsarczyk_treeshap_2021} that is used for
this work. \citet{molnar2022} provides good explanations of the
different methods and how to apply them to different models.
